---
title: "MANOVA for Trabecular Bone MicroCT Data Analysis"
author: "TF"
date: "2025-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# --- 1. Install and load necessary packages ---

'dplyr' is used for data manipulation. 'rstatix' provides a user-friendly interface for statistical tests. 'ggpubr' is great for data visualization.

## Check if packages are installed, if not, install them

```{r}
if(!require(dplyr)) install.packages("dplyr")
if(!require(rstatix)) install.packages("rstatix")
if(!require(ggpubr)) install.packages("ggpubr")
```

## Load the packages

```{r}
library(dplyr)
library(rstatix)
library(ggpubr)
```

# --- 2. Create or load your data ---

Each row is a sample (e.g., one mouse tibia). 'Group' column indicates Control (Con) or Experiment (Exp). Other columns are your trabecular bone parameters.

## mG female OVX data - tibia trabecular bone

```{r}
my_data <- data.frame(
  Group = rep(c("Con", "Exp"), each = 5),
  Tb.BV.TV = c(14.53, 14.16, 16.2, 17.45, 16.43,14.13,14.94,13.77,10.39,11.03), 
  Tb.N = c(4.14,4.67,4.48,4.83,4.22,3.99,3.85,3.71,3.29,3.41),
  Tb.BMD = c(174.391323,180.70556,180.098287,184.629318,174.102078,174.491972,174.125246,171.085229,150.253229,153.522325)
)
```

## mG female OVX data - tibia cortical bone

```{r}
my_data <- data.frame(
  Group = rep(c("Con", "Exp"), each = 5),
  Ct.BA.TA = c(46.33,43.25,46.7,46.29,46.87,40.77,42.12,38.95,28.97,32.09), 
  Ct.BMD = c(1805.48,1759.67,1791.86,1837.26,1856.18,1776.8,1782.42,1733.28,1624.32,1641.45),
  Ct.Th = c(0.2,0.18,0.21,0.22,0.22,0.18,0.19,0.18,0.14,0.15)
)
```

# --- 3. Data Transformation (Optional but Recommended) ---

Standardizing the data (scaling) can improve the interpretation of some post-hoc analyses and diagnostic plots. It centers the data around 0 with a standard deviation of 1.

```{r}
scaled_data <- my_data %>%
  # mutate_at(vars(BV.TV, Tb.N, Tb.Th, SMI), scale)
  # mutate_at(vars(Tb.BV.TV, Tb.N, Tb.BMD), scale)
  mutate_at(vars(Ct.BA.TA, Ct.BMD, Ct.Th), scale)
```

# --- 4. Perform MANOVA ---

We want to see if the 'Group' has a significant effect on the combination of all trabecular bone parameters.

The formula `cbind(BV.TV, Tb.N, Tb.Th, SMI) ~ Group` means we are testing the effect of 'Group' on the set of variables.

```{r}
# manova_result <- manova(cbind(BV.TV, Tb.N, Tb.Th, SMI) ~ Group, data = my_data)
# manova_result <- manova(cbind(Tb.BV.TV, Tb.N, Tb.BMD) ~ Group, data = my_data)
manova_result <- manova(cbind(Ct.BA.TA, Ct.BMD, Ct.Th) ~ Group, data = my_data)
```

Print the summary of the MANOVA results We look at the Pr(\>F) value (p-value). If it's \< 0.05, it indicates a statistically significant effect of the 'Group' on the trabecular bone parameters as a whole.

```{r}
summary(manova_result, test = "Pillai") # Pillai's trace is a robust test statistic
```

# --- 5. Post-hoc Analysis (if MANOVA is significant) ---

If the overall MANOVA result is significant, you can investigate which specific parameters are different between the groups. You can perform separate t-tests or ANOVAs for each parameter. It is good practice to adjust p-values for multiple comparisons.

Using the 'rstatix' package for easy post-hoc tests with p-value adjustment

```{r}
post_hoc_tests <- my_data %>%
  gather(key = "Parameter", value = "Value", -Group) %>%
  group_by(Parameter) %>%
  t_test(Value ~ Group) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance()

# Print the results of the post-hoc tests
print(post_hoc_tests)
```

# --- 6. Visualization ---

Visualizing the data is always a good idea. Here is an example using boxplots for each parameter.

Reshape data for plotting with ggplot2

```{r}
plot_data <- my_data %>%
  gather(key = "Parameter", value = "Value", -Group)
```

Create the plot

```{r}
ggboxplot(
  plot_data, x = "Group", y = "Value",
  color = "Group", palette = "jco",
  facet.by = "Parameter", scales = "free_y" # 'free_y' is important because of different scales
) +
  stat_compare_means(
    aes(label = ..p.signif..),
    method = "t.test", label.x = 1.5
  ) +
  labs(
    title = "Trabecular Bone Parameters: Control vs. Experiment",
    x = "Experimental Group",
    y = "Parameter Value"
  )
```
